Hybrid Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-19 manuscript/main.tex
previous literature separated

	%Infinite dimensional beta estimation problem (ill-posed problem)
	%Multicollinearity and high-dimensionality in scalar predictors are not handled.
	%Also, there may be high correlation between functional and scalar predictors.
	This hybrid model poses several challenges. 
	First, estimating the infinite-dimensional slope function $\beta_k(t)$ from a finite number of data
	is an ill-posed problem.% that requires dimension reduction or regularization based on structural assumptions, such as derivatives and covariance patterns.
	Second, we face the high-dimensionality of the predictors with $K+p$ dimensions.
	Third, strong correlations between functional and scalar predictors are common but often overlooked by separate modeling approaches.

	To address the ill-posedness stemming from the infinite-dimensionality of the functional components, common remedies include basis expansion,
	using power-series  \citep{goldsmithPenalizedFunctionalRegression2011},
	B-splines \citep{cardotSplineEstimatorsFunctional2003, caiPredictionFunctionalLinear2006}, or 
	wavelets \citep{zhaoWaveletbasedLASSOFunctional2012},
	and structure-aware regularization, such as roughness penalties \jongmin{let's add citations here}.
	For the high-dimensionality challenge, a major solution is regression with derived inputs,
	such as Principal Component Analysis (PCA) regression.
	This involves applying (multivariate) functional PCA (FPCA)  to the functional predictors, possibly with standard PCA to the scalar predictors,
	followed by a classical multivariate regression on the resulting scores.
	This PCA regression strategy is well-studied for functional predictors alone \citep{hallMethodologyConvergenceRates2007, reissFunctionalPrincipalComponent2007, febrero-bandeFunctionalPrincipalComponent2017}.
	However, this approach has a major drawback: the derived inputs are not informed by the response variable. Consequently, they are not guaranteed to capture the core regression relationship with a small number of derived inputs.
	Partial Least Squares (PLS) regression offers a powerful alternative by explicitly seeking low-dimensional projections of the predictors that maximize their covariance with the outcome. PLS has been extensively studied in the context of functional data, both for single and multiple functional predictors \citep{predaPLSRegressionStochastic2005, reissFunctionalPrincipalComponent2007, aguileraUsingBasisExpansions2010, delaigleMethodologyTheoryPartial2012, aguileraPenalizedVersionsFunctional2016, febrero-bandeFunctionalPrincipalComponent2017, beyaztasRobustFunctionalPartial2022, saricamPartialLeastsquaresEstimation2022, mutisFunctiononfunctionLinearQuantile2025}.
	However, all of these existing PCA or PLS approaches overlook the potentially strong correlations between functional and scalar components. This can lead to multicollinearity and suboptimal predictive performance.
	To address the third challenge, \citet{jangPrincipalComponentAnalysis2021}  proposed a joint PCA method that accounts for correlations between functional and scalar data. However, the resulting components are still not informed by the response, so they may fail to capture the directions of maximal correlation with the outcome.

>>>>

	%Infinite dimensional beta estimation problem (ill-posed problem)
	%Multicollinearity and high-dimensionality in scalar predictors are not handled.
	%Also, there may be high correlation between functional and scalar predictors.
	This hybrid model poses several challenges. 
	First, estimating the infinite-dimensional slope function $\beta_k(t)$ from a finite number of data
	is an ill-posed problem.% that requires dimension reduction or regularization based on structural assumptions, such as derivatives and covariance patterns.
	Second, we face the high-dimensionality of the predictors with $K+p$ dimensions.
	Third, strong correlations between functional and scalar predictors are common but often overlooked by separate modeling approaches.
	This paper unifies the remedies proposed for these three issues
by proposing a hybrid partial least square (PLS) regression framework that integrates functional and vector predictors in a principled and coherent manner. Before we summarize the key contributions of this new framework, we will first discuss previous approaches and their limitations.

	\subsection{Previous works and limitations}
	To address the ill-posedness stemming from the infinite-dimensionality of the functional components, common remedies include basis expansion,
	using power-series  \citep{goldsmithPenalizedFunctionalRegression2011},
	B-splines \citep{cardotSplineEstimatorsFunctional2003, caiPredictionFunctionalLinear2006}, or 
	wavelets \citep{zhaoWaveletbasedLASSOFunctional2012},
	and structure-aware regularization, such as roughness penalties \jongmin{let's add citations here}.
	For the high-dimensionality challenge, a major solution is regression with derived inputs,
	such as Principal Component Analysis (PCA) regression.
	This involves applying (multivariate) functional PCA (FPCA)  to the functional predictors, possibly with standard PCA to the scalar predictors,
	followed by a classical multivariate regression on the resulting scores.
	This PCA regression strategy is well-studied for functional predictors alone \citep{hallMethodologyConvergenceRates2007, reissFunctionalPrincipalComponent2007, febrero-bandeFunctionalPrincipalComponent2017}.
	However, this approach has a major drawback: the derived inputs are not informed by the response variable. Consequently, they are not guaranteed to capture the core regression relationship with a small number of derived inputs.
	Partial Least Squares (PLS) regression offers a powerful alternative by explicitly seeking low-dimensional projections of the predictors that maximize their covariance with the outcome. PLS has been extensively studied in the context of functional data, both for single and multiple functional predictors \citep{predaPLSRegressionStochastic2005, reissFunctionalPrincipalComponent2007, aguileraUsingBasisExpansions2010, delaigleMethodologyTheoryPartial2012, aguileraPenalizedVersionsFunctional2016, febrero-bandeFunctionalPrincipalComponent2017, beyaztasRobustFunctionalPartial2022, saricamPartialLeastsquaresEstimation2022, mutisFunctiononfunctionLinearQuantile2025}.
	However, all of these existing PCA or PLS approaches overlook the potentially strong correlations between functional and scalar components. This can lead to multicollinearity and suboptimal predictive performance.
	To address the third challenge, \citet{jangPrincipalComponentAnalysis2021}  proposed a joint PCA method that accounts for correlations between functional and scalar data. However, the resulting components are still not informed by the response, so they may fail to capture the directions of maximal correlation with the outcome.
